---
title: "EC233 Homework 1"
date: "01.11.2024"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: united
    code_folding: hide
    fig_caption: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  fig.align = 'center'
)
```

# Setup and Package Installation

## Required Packages

```{r install_packages}
packages <- c("devtools", "quantmod", "ggplot2", "dplyr", "zoo")
invisible(lapply(packages, function(p) {
  if (!require(p, character.only = TRUE)) {
    install.packages(p)
    library(p, character.only = TRUE)
  }
}))

if (!require("pwt10")) {
  devtools::install_github("michaelchleung/pwt10")
  library(pwt10)
}
```

# Creating Dataset {.tabset}

## Statistics

```{r creating_dataset}
set.seed(7623)
n <- sample(35:4000, 1)
mu <- rnorm(1, 2.5, 4)
sd <- abs(rnorm(1, 2.5, 4))

# Generate datasets
ulusoy_data <- rnorm(n, mu, sd)
ulusoy_data2 <- rnorm(n, mu, 2.5*sd)

stats <- data.frame(
  Statistic = c("Mean", "Variance", "Standard Deviation", "Median"),
  Value = round(c(mean(ulusoy_data), var(ulusoy_data), 
                 sd(ulusoy_data), median(ulusoy_data)), 4)
)

knitr::kable(stats, caption = "Statistics")
```

## Histograms

```{r histograms}
par(mar = c(5, 4, 4, 4))
hist(ulusoy_data, 
     breaks = 30,
     probability = TRUE, 
     col = rgb(0,0,1,0.5),
     main = "Histograms",
     xlab = "Values")

hist(ulusoy_data2, 
     breaks = 30,
     probability = TRUE,
     col = rgb(1,0,0,0.4),
     add = TRUE)

legend("topright", 
       c("Original Data", "Data with 2.5*SD"), 
       fill = c(rgb(0,0,1,0.5), rgb(1,0,0,0.5)))
cat("Since standard deviation measures how spread out numbers are from the mean, higher the standard deviation, higher the data, hence a broader histogram.")
```

# Macro-economic Analysis {.tabset}

## GDP Comparisons of 2 Sets of Countries

### Code
```{r gdp_comparisons}
gdp_compare <- function(data, c1, c2, year1, year2, type = "growth") {
  if (type == "growth") {
    comparison_data <- data %>%
      filter(isocode %in% c(c1, c2), 
             year >= year1, 
             year <= year2) %>%
      group_by(isocode) %>%
      arrange(year) %>%
      mutate(growth = c(NA, diff(log(rgdpna/pop))) * 100) %>%
      filter(!is.na(growth))
    
    plot <- ggplot(comparison_data, 
           aes(x = year, y = growth, color = isocode)) +
      geom_line() +
      theme_minimal() +
      labs(title = paste("GDP per Capita Growth Rate:", c1, "vs", c2),
           x = "Year", 
           y = "Growth Rate (%)")
  } else {
    comparison_data <- data %>%
      filter(isocode %in% c(c1, c2), 
             year >= year1, 
             year <= year2) %>%
      group_by(isocode) %>%
      arrange(year) %>%
      mutate(nominal_gdp = rgdpna) %>%
      filter(!is.na(nominal_gdp))
    
    plot <- ggplot(comparison_data, 
           aes(x = year, y = nominal_gdp, color = isocode)) +
      geom_line() +
      theme_minimal() +
      labs(title = paste("Nominal GDP Comparison:", c1, "vs", c2),
           x = "Year", 
           y = "GDP (in millions)")
  }
  return(plot)
}
```

### Turkey vs Poland
```{r turkey_poland}
data("pwt10.0")
print(gdp_compare(pwt10.0, "TUR", "POL", 1987, 2015, type = "growth"))
```

### Germany vs France
```{r germany_france}
print(gdp_compare(pwt10.0, "DEU", "FRA", 1987, 2015, type = "nominal"))
```

## Turkey's Ranking Among 10 Similar Countries

### Code
```{r countries_analyzing}
analyze_countries <- function(data, countries, year1, year2) {
  rankings <- data %>%
    filter(isocode %in% countries, 
           year >= year1, 
           year <= year2) %>%
    group_by(isocode) %>%
    arrange(year) %>%
    mutate(
      gdp_pc = rgdpna/pop,
      growth = c(NA, diff(log(gdp_pc))) * 100
    ) %>%
    group_by(year) %>%
    mutate(rank = rank(-growth, ties.method = "min")) %>%
    ungroup() %>%
    filter(!is.na(growth))
  
  p1 <- ggplot(rankings, aes(x = year, y = growth, color = isocode)) +
    geom_line() +
    theme_minimal() +
    labs(title = "Comparing GDP per Capita Growth Rates for Selected Countries",
         x = "Year", 
         y = "Growth Rate (%)",
         color = "Country") +
    theme(legend.position = "bottom")
  
turkey_ranks <- rankings %>%
  filter(isocode == "TUR") %>%
  select(year, rank)

p2 <- ggplot(turkey_ranks, aes(x = year, y = rank)) +
  geom_line(color = "blue") +
  geom_point(color = "red") +
  scale_y_reverse(breaks = 1:11) +
  theme_minimal() +
  labs(title = "Turkey's Ranking",
       x = "Year",
       y = "Rank",
       caption = "Countries compared with: Germany, Russia, Indonesia, Italy, Pakistan\nSouth Korea, Greece, Egypt, Hungary, Argentina") +
  theme(plot.caption = element_text(hjust = 0.5, margin = margin(t = 10)))
  
  return(list(growth_plot = p1, rank_plot = p2))
}
```


### Comparing Growth Rates
```{r ten_similar_countries}
countries <- c("TUR", "DEU", "RUS", "IDN", "ITA", "PAK", 
              "KOR", "GRC", "EGY", "HUN", "ARG")
plots <- analyze_countries(pwt10.0, countries, 1987, 2015)
print(plots$growth_plot)
```

### Turkey's Ranking
```{r turkey_ranking}
print(plots$rank_plot)
```

# Stock Market {.tabset}

```{r stock_market}
safe_get_stock <- function(symbol, date1, date2) {
  tryCatch({
    getSymbols(symbol, 
               src = "yahoo", 
               from = date1, 
               to = date2, 
               periodicity = "weekly",
               auto.assign = FALSE)
  }, error = function(e) {
    message(paste("Error for", symbol, ":", e$message))
    return(NULL)
  })
}

analyze_stocks <- function(symbols, date1, date2) {
  results <- list()
  combined_data <- list()
  
  for(symbol in symbols) {
    stock_data <- safe_get_stock(symbol, date1, date2)
    
    if(is.null(stock_data)) {
      message(paste("Error"))
      next
    }
    
    prices <- try(Cl(stock_data))
    returns <- try({
      ret <- diff(log(prices)) * 100
      ret[!is.finite(ret)] <- NA
      ret
    })
    roll_sd <- try(rollapply(returns, 8, sd, fill = NA))
    
    combined_data[[symbol]] <- list(
      prices = data.frame(
        date = index(prices),
        price = as.numeric(prices)
      ),
      returns = data.frame(
        date = index(returns),
        returns = as.numeric(returns),
        roll_sd = as.numeric(roll_sd)
      )
    )
    
    results[[symbol]] <- list(
      stats = try({
        c(
          Max = max(prices, na.rm = TRUE),
          Min = min(prices, na.rm = TRUE),
          Range = diff(range(prices, na.rm = TRUE)),
          Average = mean(prices, na.rm = TRUE),
          SD = sd(as.numeric(prices), na.rm = TRUE),
          CV = sd(as.numeric(prices), na.rm = TRUE) / 
              mean(as.numeric(prices), na.rm = TRUE)
        )
      }),
      corr = cor(returns, roll_sd, use = "complete.obs")
    )
  }
  
  # 1. Combined price comparison
  price_data <- merge(combined_data[[symbols[1]]]$prices, 
                     combined_data[[symbols[2]]]$prices, 
                     by = "date", suffixes = c("_1", "_2"))
  
  price_plot <- ggplot(price_data, aes(x = date)) +
    geom_line(aes(y = price_1, color = symbols[1])) +
    geom_line(aes(y = price_2, color = symbols[2])) +
    theme_minimal() +
    labs(title = "Stock Price Comparison",
         x = "Date", 
         y = "Price",
         color = "Stock")
  
  stats_text <- paste0(
    symbols[1], " Statistics:\n",
    "Max: ", round(results[[symbols[1]]]$stats["Max"], 2),
    ", Min: ", round(results[[symbols[1]]]$stats["Min"], 2),
    ", Range: ", round(results[[symbols[1]]]$stats["Range"], 2),
    ", Average: ", round(results[[symbols[1]]]$stats["Average"], 2),
    ", SD: ", round(results[[symbols[1]]]$stats["SD"], 2),
    ", CV: ", round(results[[symbols[1]]]$stats["CV"], 4), "\n\n",
    symbols[2], " Statistics:\n",
    "Max: ", round(results[[symbols[2]]]$stats["Max"], 2),
    ", Min: ", round(results[[symbols[2]]]$stats["Min"], 2),
    ", Range: ", round(results[[symbols[2]]]$stats["Range"], 2),
    ", Average: ", round(results[[symbols[2]]]$stats["Average"], 2),
    ", SD: ", round(results[[symbols[2]]]$stats["SD"], 2),
    ", CV: ", round(results[[symbols[2]]]$stats["CV"], 4), "\n\n",
    "Risk Analysis: ", 
    ifelse(results[[symbols[1]]]$stats["CV"] > results[[symbols[2]]]$stats["CV"],
           paste(symbols[1], "is riskier than", symbols[2]),
           paste(symbols[2], "is riskier than", symbols[1])),
    " because of higher Coefficient of Variation (CV), which means greater standard deviation (SD). As data spreads more out, \nit becomes riskier and volatile."
  )
  
  price_plot <- price_plot + 
    labs(caption = stats_text) +
    theme(plot.caption = element_text(hjust = 0))
  
  returns_data <- merge(combined_data[[symbols[1]]]$returns,
                       combined_data[[symbols[2]]]$returns,
                       by = "date", suffixes = c("_1", "_2"))
  
  returns_plot <- ggplot(returns_data, aes(x = date)) +
    geom_line(aes(y = returns_1, color = symbols[1])) +
    geom_line(aes(y = returns_2, color = symbols[2])) +
    theme_minimal() +
    labs(title = "Comparing Weekly Returns",
         x = "Date", 
         y = "Returns (%)",
         color = "Stock")
  
  volatility_plot <- ggplot(returns_data, aes(x = date)) +
    geom_line(aes(y = roll_sd_1, color = symbols[1])) +
    geom_line(aes(y = roll_sd_2, color = symbols[2])) +
    theme_minimal() +
    labs(title = "8 Week Rolling Standard Deviation",
         x = "Date", 
         y = "Standard Deviation",
         color = "Stock",
         caption = paste("Volatility patterns indicate temporal clustering,",
                        "showing periods of high and low market stress",
                        "having similar impacts on both stocks."))
  
  returns_vol_plots <- list()
  for(i in 1:2) {
    symbol <- symbols[i]
    suffix <- paste0("_", i)
    
    returns_vol_plots[[symbol]] <- ggplot(returns_data, aes(x = date)) +
      geom_line(aes(y = get(paste0("returns", suffix)), color = "Returns")) +
      geom_line(aes(y = get(paste0("roll_sd", suffix)), color = "Volatility")) +
      theme_minimal() +
      labs(title = paste(symbol, "Returns vs Volatility"),
           x = "Date",
           y = "Value",
           color = "Metric",
           caption = paste("Correlation coefficient:", 
                         round(results[[symbol]]$corr, 3),
                         "\nNegative correlation shows",
                         "periods of high losses",
                         "usually come with great volatility."))
  }
  
  return(list(
    price_plot = price_plot,
    returns_plot = returns_plot,
    volatility_plot = volatility_plot,
    returns_vol_plots = returns_vol_plots,
    statistics = results
  ))
}

# Usage example:
date_start <- as.Date("2005-05-05")
date_end <- as.Date("2023-10-01")
analysis <- analyze_stocks(c("MSFT", "AAPL"), date_start, date_end)

# Display plots
print(analysis$price_plot)
print(analysis$returns_plot)
print(analysis$volatility_plot)
print(analysis$returns_vol_plots$MSFT)
print(analysis$returns_vol_plots$AAPL)

date_start <- as.Date("2005-05-05")
date_end <- Sys.Date() - 1
stocks <- analyze_stocks(c("MSFT", "AAPL"), date_start, date_end)

if(!is.null(stocks$MSFT)) {
  print(stocks$MSFT$prices)
}

if(!is.null(stocks$MSFT)) {
  print(stocks$MSFT$returns)
}

if(!is.null(stocks$MSFT) && !inherits(stocks$MSFT$stats, "try-error")) {
  knitr::kable(data.frame(
    Statistic = names(stocks$MSFT$stats),
    Value = round(stocks$MSFT$stats, 2)
  ), caption = "Microsoft Statistics")
}

if(!is.null(stocks$AAPL)) {
  print(stocks$AAPL$prices)
}

if(!is.null(stocks$AAPL)) {
  print(stocks$AAPL$returns)
}

if(!is.null(stocks$AAPL) && !inherits(stocks$AAPL$stats, "try-error")) {
  knitr::kable(data.frame(
    Statistic = names(stocks$AAPL$stats),
    Value = round(stocks$AAPL$stats, 2)
  ), caption = "Apple Statistics")
}
```

# Probability Calculations {.tabset}

## Birthday Problem

```{r birthday}
birthday_probability <- function(n, x) {
  if(x > n) return(0)
  
  n_simulations <- 1000
  successes <- 0
  
  for(i in 1:n_simulations) {
    birthdays <- sample(1:365, n, replace = TRUE)
    counts <- table(birthdays)
    successes <- successes + (max(counts) >= x)
  }
  
  return(successes / n_simulations)
}

n_values <- seq(1, 555, by = 10)
x_values <- 2:5

results <- data.frame(n = n_values)
for(x in x_values) {
  probs <- sapply(n_values, function(n) birthday_probability(n, x))
  results[paste0("x", x)] <- probs
}

library(ggplot2)
library(gridExtra)

# Plot for x = 2
p1 <- ggplot(results, aes(x = n, y = x2)) +
  geom_line(color = "pink", size = 1) +
  theme_minimal() +
  labs(title = "At least 2 people sharing birthday",
       x = "Class Size (n)",
       y = "Probability") +
  ylim(0, 1) +
  theme(plot.title = element_text(size = 10))

# Plot for x = 3
p2 <- ggplot(results, aes(x = n, y = x3)) +
  geom_line(color = "red", size = 1) +
  theme_minimal() +
  labs(title = "At least 3 people sharing birthday",
       x = "Class Size (n)",
       y = "Probability") +
  ylim(0, 1) +
  theme(plot.title = element_text(size = 10))

# Plot for x = 4
p3 <- ggplot(results, aes(x = n, y = x4)) +
  geom_line(color = "turquoise", size = 1) +
  theme_minimal() +
  labs(title = "At least 4 people sharing birthday",
       x = "Class Size (n)",
       y = "Probability") +
  ylim(0, 1) +
  theme(plot.title = element_text(size = 10))

# Plot for x = 5
p4 <- ggplot(results, aes(x = n, y = x5)) +
  geom_line(color = "yellow", size = 1) +
  theme_minimal() +
  labs(title = "At least 5 people sharing birthday",
       x = "Class Size (n)",
       y = "Probability") +
  ylim(0, 1) +
  theme(plot.title = element_text(size = 10))

# Combine all plots in a 2x2 grid
grid.arrange(p1, p2, p3, p4, ncol = 2)

cat("Had problems with generating a 3d graph, therefore found a solution this way.")
```

## Coin Toss
```{r coin_toss}
simulate_coin_tosses <- function(p_heads = 0.5) {
  N <- 500
  results <- numeric(N)
  cumulative_prob <- numeric(N)
  
  for(i in 1:N) {
    tosses <- rbinom(4, 1, p_heads)
    results[i] <- sum(tosses) >= 1
    cumulative_prob[i] <- mean(results[1:i])
  }
  
  return(data.frame(
    experiment = 1:N,
    cumulative_probability = cumulative_prob
  ))
}

# Running simulations
set.seed(123)
fair_coin_results <- simulate_coin_tosses(0.5)
unfair_coin_results <- simulate_coin_tosses(0.2)

# Fair coin
fair_coin_plot <- ggplot(fair_coin_results, aes(x = experiment, y = cumulative_probability)) +
  geom_line(color = "green") +
  geom_hline(yintercept = 1 - (1/2)^4, linetype = "dashed", color = "red") +
  theme_minimal() +
  labs(title = "Fair Coin (p = 0.5): Probability of At Least One Head in Four Tosses",
       x = "Number of Experiments",
       y = "Probability",
       caption = "Red line is the theoretical probability = 0.9375") +
  theme(text = element_text(size = 12),
        plot.caption = element_text(size = 12)) +
  ylim(0, 1)

# Unfair coin
unfair_coin_plot <- ggplot(unfair_coin_results, aes(x = experiment, y = cumulative_probability)) +
  geom_line(color = "purple") +
  geom_hline(yintercept = 1 - (0.8)^4, linetype = "dashed", color = "blue") +
  theme_minimal() +
  labs(title = "Unfair Coin (p = 0.2): Probability of At Least One Head in Four Tosses",
       x = "Number of Experiments",
       y = "Probability",
       caption = "Blue line is the theoretical probability = 0.5904") +
  theme(text = element_text(size = 12),
        plot.caption = element_text(size = 12)) +
  ylim(0, 1)

print(fair_coin_plot)
print(unfair_coin_plot)
```

## Elections
```{r election}
simulate_election <- function(n_voters, p_A, p_B, p_C, n_trials = 500) {
  sample_sizes <- c(5, seq(10, n_voters, by = 10))
  
  win_matrix <- matrix(0, nrow = length(sample_sizes), ncol = 3)
  colnames(win_matrix) <- c("A", "B", "C")
  
  for(i in seq_along(sample_sizes)) {
    current_n <- sample_sizes[i]
    winners <- character(n_trials)
    
    for(j in 1:n_trials) {
      votes <- sample(c("A", "B", "C"), current_n, 
                     prob = c(p_A, p_B, p_C), 
                     replace = TRUE)
      
      vote_counts <- table(factor(votes, levels = c("A", "B", "C")))
      
      # Find out winner
      max_votes <- max(vote_counts)
      winners[j] <- sample(names(vote_counts)[vote_counts == max_votes], 1)
    }
    
    win_props <- table(factor(winners, levels = c("A", "B", "C"))) / n_trials
    win_matrix[i,] <- win_props
  }
  
  # Create data frame
  results_df <- data.frame(
    sample_size = rep(sample_sizes, 3),
    win_probability = c(win_matrix[,"A"], win_matrix[,"B"], win_matrix[,"C"]),
    candidate = rep(c("A", "B", "C"), each = length(sample_sizes))
  )
  
  return(results_df)
}

# Run simulations for both scenarios
set.seed(123)

# Scenario 1: 0.55, 0.25, 0.20
results1 <- simulate_election(555, 0.55, 0.25, 0.20)

plot1 <- ggplot(results1, aes(x = sample_size, y = win_probability, color = candidate)) +
  geom_line() +
  geom_hline(yintercept = 0.55, linetype = "dashed", color = "orange", alpha = 0.5) +
  geom_hline(yintercept = 0.25, linetype = "dashed", color = "green", alpha = 0.5) +
  geom_hline(yintercept = 0.20, linetype = "dashed", color = "blue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Convergence of Winning Probabilities (Voting Ratios: A=0.55, B=0.25, C=0.20)",
       x = "Sample Size (n)",
       y = "Winning Probability",
       color = "Candidate") +
  ylim(0, 1)

# Scenario 2: 0.42, 0.48, 0.1
results2 <- simulate_election(555, 0.42, 0.48, 0.10)

plot2 <- ggplot(results2, aes(x = sample_size, y = win_probability, color = candidate)) +
  geom_line() +
  geom_hline(yintercept = 0.42, linetype = "dashed", color = "orange", alpha = 0.5) +
  geom_hline(yintercept = 0.48, linetype = "dashed", color = "green", alpha = 0.5) +
  geom_hline(yintercept = 0.1, linetype = "dashed", color = "blue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Convergence of Winning Probabilities (Voting Ratios: A=0.42, B=0.48, C=0.10)",
       x = "Sample Size (n)",
       y = "Winning Probability",
       color = "Candidate") +
  ylim(0, 1)

print(plot1)
print(plot2)

# Analysis
cat("\nAnalysis of Election Winning Probabilities:\n\n")
cat("1. Scenario 1 (0.55, 0.25, 0.20):\n")
cat("   - Winning probabilities don't converge to voting ratios\n")
cat("   - Candidate A wins much more frequently than his voting ratio\n")
cat("2. Scenario 2 (0.42, 0.48, 0.1):\n")
cat("   - In the 2nd scenerio I tried to simulate a probable Turkish election\n")
cat("   - There is a competition between Reis and Primary CHP Candidate\n")
cat("   - Even though it is a low one, REİS still has a chance\n")
cat("   - DEM has no chance of winning despite a 10% vote share\n")
cat("   - This demonstrates how plurality voting can affect minorities disportionately\n\n")
cat("3. Additional Analysis Possibilities:\n")
cat("   - Majority or Plurality Trends Over Time\n")
cat("   - Sensitivity Analysis\n")
cat("   - Polarization Analysis in Highly Polarized Countries\n")
cat("   - Coalition scenarios\n")

```